# Weather Data - Cutting Room 

## Download hourly CIMIS data via the API. 

This material has been dropped because the API only lets you download 1750 records at a time (so you can only get a month or two of hourly data at a time).

The CIMIS FTP server is also currently not working.

A better method is generating the CSV file from the website

Step 1. Define the CSV file where we'll save the data:

```{r}
cim_ply_hr_fn <- "./data/cim_ply_hr.csv"
```

```{r eval = file.exists(cim_ply_hr_fn)}
## If the local csv file already exists, then just quietly import it.
## This prevents us from having to call CIMIS every time the ebook is rendered.
#cim_ply_dly_tbl <- read_csv(cim_ply_dly_fn, col_types = "Dddd")
```

\

Hourly CIMIS data via the API. 

::: {.callout-note}
### CIMIS API Limit

The CIMIS API limits the number of records you can retun to 1750. To get hourly data for three variables for an entire year, we need to grab them month-by-month.

:::

\


```{r eval = FALSE}
## Set your CIMIS key - already done (see above)


month_num <- 11

start_dt <- make_date(year = 2023, month = month_num, day = 1)
start_dt

end_dt <- make_date(year = 2023, month = month_num + 1, day = 1) - 1
end_dt

### Query CIMIS data (Plymouth station #227)
cimis_plymouth_hourly_tbl <- cimis_data(targets = 227, 
                                        start.date = "2023-01-01", 
                                        end.date = "2023-01-31",
                                        measure.unit = "E",
                                        items = "hly-air-tmp,hly-precip")


cim_ply_lng_tbl <- cimis_data(targets = 227, 
                         start.date = "2023-01-01", 
                         end.date = "2023-12-31",
                         measure.unit = "E",
                         items = "day-precip,day-air-tmp-max,day-air-tmp-min")

cim_ply_dly_tbl <- cim_ply_lng_tbl |> 
  select(Date, Item, Value) |> 
  pivot_wider(id_cols = Date, names_from = Item, values_from = Value) |> 
  rename(date = Date, tmax_f = DayAirTmpMax, tmin_f = DayAirTmpMin, prec = DayPrecip)

## Save it to disk
write_csv(cim_ply_dly_tbl, file = cim_ply_dly_fn)
```


```{r eval = FALSE}
## Define where to save this
cim_ply_hr_fn <- "./data/cim_ply_hr.csv"

if (file.exists(cim_ply_hr_fn)) {
  
  

} else {
  ## Step 1. Load my cimis key
  my_cimis_key <- readLines("~/My Keys/cimis_webapi.txt", n=1)
  cimir::set_key(my_cimis_key)

  
  ## NEED TO FINISH
  ##  - CIMIS API only allows 1750 records, so to get 1 year of data we need to loop thru the months
  ##    (or download from the web)
  
  # cim_ply_hr_tbl <- cim_ply_hrlng_tbl |> 
  #   select(Date, Item, Value) |> 
  #   pivot_wider(id_cols = Date, names_from = Item, values_from = Value) |> 
  #   rename(date = Date, tmax = DayAirTmpMax, tmin = DayAirTmpMin, prec = DayPrecip)
  # 
  # write_csv(cim_ply_dly_tbl, file = cim_ply_dly_fn)
  
}
```
